Asymmetric relationships and their effects on coexistence

### 2023-02-28 paper/Paper.Rnw
took another look at the Results I edited the other day and made some further presentational improvements

%the beta fucundities version
The \emph{beta fecundities lottery model}
used $B_i = \eta_i F_\beta^{-1} \circ \varphi(b_i)$, where $\varphi$ is the CDF of the standard normal
distribution and $F_\beta$ is the CDF of a beta distribution 
with shape and scale parameters $0.5$ (probability density function proportional to $x^{-05}(1-x)^{-0.5}$ on
the unit interval). So the beta fecundities model had parameters $\eta_1$, $\eta_2$ and $\delta$,
and produced fecundities with $U$-shaped, beta distributions between $0$ and $\eta_1$, for $B_1$, and
$\eta_2$, for $B_2$. The mean fecundity for species $i$ was $\eta_i/2$, and larger $\eta_i$ also 
corresponded to more variable fecundities for species $i$. The parameter $\eta_i$ represents the upper
bound on what fecundities could occur for species $i$.

@@

For each combination of $\delta=0.2, 0.4, 0.6$ and $\mu_1-\mu_2=0,-0.5,-2,-4$,
we plotted the components of (\ref{eq:full_decomp}) against $\sigma$ for $\sigma$ ranging 
from $0$ to $7$. Some of 
the $\mu$ differences and $\sigma$ values we considered were very unrealistic (see Results),
so we considered after plotting whether conclusions about the importance of ATAs depended 
on realistic values of parameters.
%***DAN: Jasmin mentioned that she only used \delta=0.2, .4, .6 because of a comment Ellner made where he 
%said it was known that larger values of \delta made for very little storage effects. Read that comment 
%again and then write something here to jstify the choice. I think I can probably then delete the section of
%sup mat, mentioned in the commented text below, that talks about \delta=1.
%Mathematical simplications for the $\delta=1$ case make it straightforward to see that
%coexistence is impossible and that the case is relatively trivial (SI 
%section \ref{SI-sect:delta1}), so that case is not plotted.  
The simulation methods of \cite{Ellner_2019} were used to compute the components of (\ref{eq:full_decomp}).
Computationally efficient means of performing these simulations are in 
SI section \ref{SI-sect:efficient_computation}.

@@

For each combination of $\delta=0.2, 0.4, 0.6$
and for left- and right-tail associated noise, we plotted 
the components of (\ref{eq:full_decomp}) against the upper bound ratio $\eta_2/\eta_1$, using 
the fairly realistic range $1 \leq \eta_2/\eta_1 \leq 5$. 
The simulation methods of \cite{Ellner_2019} were again used.
Computationally efficient means of performing these simulations are in 
SI section \ref{SI-sect:efficient_computation}.

@@

%G2 
To begin fulfilling goal G2 of the Introduction (to explore, using the lottery model, the relative magnitudes 
of the contributions to coexistence that come from ATAs compared to other mechanisms; and to see if ATA rescue 
or ATA exclusion can occur), Fig.
\ref{fig:fig3} shows that, for the log-normal fecundities lottery model, 
the ATA term of (\ref{eq:full_decomp}) sometimes, but not always, has 
magnitude comparable to the magnitudes of other terms of the extended MCT decomposition. 
Magnitudes of $\Delta_i^{[EC]}$ tended to be larger, relative to the magnitudes of the 
other terms of (\ref{eq:full_decomp}), for smaller-magnitude differences between $\mu_1$ and 
$\mu_2$, i.e., for closer-to-equal competitors. 
In the $\mu_1=\mu_2$ case (panels a-c), for which the two species were equal 
competitors, storage effects ($\Delta_i^{[EC]}+\Delta_i^{[E\nat C]}$) 
were the only means by which coexistence could occur, and ATA effects
tended to be negative, weakening total storage effects. The ATA term 
was also often comparable in magnitude to the other terms of 
the decomposition (\ref{eq:full_decomp}) for the beta fecundities lottery model 
(Fig. \ref{fig:fig4}).

Our results also show that ATAs can make the difference between coexistence 
and failure of two species to coexist: both ``ATA rescue'' and ``ATA
exclusion'' (see Theory) are possible. Starting with the log-normal fecundities 
lottery model, for some of the panels on Fig. \ref{fig:fig3}, the GWR line falls between the ATA
line and the $x$ axis for some values of $\sigma$, e.g., see panel i.
Such cases correspond to parameter combinations
for which the presence or absence of ATAs determine whether the species coexist or not. In the 
case of panel i, for values of $\sigma$ around $6$, 
species 1 is excluded with ATAs but the two species can coexist without ATAs. 
Fig. \ref{fig:fig4} highlights additional examples using the beta fecundities lottery model.

For the log-normal fecundities lottery model, the ATA term from (\ref{eq:full_decomp}) tended
only to be comparable in magnitude to the other terms, and ATA exclusion or rescue tended only to
occur, for unrealistic parameters values; but ATA terms were large and ATA rescue and exclusion occurred
for realistic parameter values for the beta fecundities lottery model. For instance, the ATA in 
Fig. \ref{fig:fig3} was fairly 
small, compared to the other terms of (\ref{eq:full_decomp}), for values of $\sigma$ less than 
about $3$. And ATA rescue and exclusion also only occurred for $\sigma$ greater than about $3$.
The standard deviation of fecundity for species 1 when 
$\mu_1=0$ and $\sigma=3$ was greater than $8000$, so values of $\sigma$ greater than $3$ are 
probably unrealistic. Thus ATAs seem to be unimportant for coexistence
for the log-normal fecundities lottery model using realistic parameter values. However, 
all the $\eta_2/\eta_1$ ratio values we plotted were fairly realistic, corresponding to
upper-bound fecundities that were up to $5$ times higher for species 2 than for species 1. And 
the ATA term plotted in Fig. \ref{fig:fig4} was comparable in magnitude 
to the other components of (\ref{eq:full_decomp}) across the whole range of $\eta_2/\eta_1$ values
considered; and both ATA rescue and exclusion occurred for values of this ratio between $1$ and $2$.
Apparently the question of whether ATAs are important for realistic parameter values depends on model specifics.

@@

%***DAN: Jasmin, this figure looks really great. Here are a few minor things I would like to ask you to
%do to further optimize:
%1) Currently there are axis labels and numbers only on the bottom and left panels, which makes sense.
%But the axis tick marks needs to occur on all panels. I noticed that this would be an imporvement
%when I found myself wanting to see where the ATA line was between the x axis and the GWR line, but it was 
%a little hard to see where the x axis was becuase there were no tick marks. It is possible to have tick marks
%but no numbers of labels in R, and that's what we need here.
%2) The font of the x axis label, "log fcundity standard deviation, \sigma" seems slightly different from the 
%other textual labels. Can you please make it consistent? Perhaps it is not bold and the others are bold, or 
%something.
\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLLN.pdf}
\caption{The contribution of ATAs to species coexistence was often meaningfully large compared to 
other components of the MCT decomposition for the log-normal fecundities lottery model.
All panels show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). For each panel, 
$\delta$ and $\mu_1-\mu_2$ 
are fixed at values specified in the margins, and GWR and components of the decomposition
are plotted against $\sigma$. The contribution of ATAs ($\Delta_i^{[EC]}$) frequently 
has magnitude comparable to other components, indicating that ATAs can contribute meaningfully
to species coexistence or failure to coexist. In some instances, the GWR line falls 
between the ATA line and the $x$ axis, indicating that, for those parameter values, ATAs 
make the difference between species
coexistence and exclusion of the weaker competitor. These instances, which are shaded, correspond to
ATA rescue or exclusion (Theory). Standard errors of plotted quantities 
were never greater than \Sexpr{round(se3,5)}, so are not displayed. This plot uses $q_{ij}=1$. An 
analogous plot (SI Fig. \ref{SI-fig:fig3_alt}) was generated 
using $q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$. Though results differed
in detail, main conclusions were the same. \label{fig:fig3}}
\end{figure}

%***DAN: Jasmin, panel labels(a, b, c, etc.) need to be added. Please use the same format for these labels
%as you do in the other figures.
\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLB.pdf}
\caption{The contribution of ATAs to species coexistence was often meaningfully large compared to 
other components of the MCT decomposition for the beta fecundities lottery model.
All panels show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). For each panel,
$\delta$ is fixed at the value specified in the upper margin, 
and GWR and components of the decomposition
are plotted against $\eta_2/\eta_1$. The contribution of ATAs ($\Delta_i^{[EC]}$) frequently 
has magnitude comparable to other components, indicating that ATAs can contribute meaningfully
to species coexistence or failure to coexist. In some instances, the GWR line falls 
between the ATA line and the $x$ axis, indicating that, for those parameter values, ATAs make 
the difference between species
coexistence and exclusion of the weaker competitor. These instances, which are shaded, correspond to
ATA rescue or exclusion (Theory). Standard errors of plotted quantities 
were never greater than \Sexpr{round(se4,5)}, so are not displayed. 
%***DAN: Jasmin, we need to replace the NA in the above line by a variable names, which your code should have saved.
%The variable needs to be loaded in the R chunk near the top of this file, and then refereced here, please.
This plot uses $q_{ij}=1$. 
%***DAN: Right, Jasmin?
An analogous plot (SI Fig. \ref{SI-fig:fig4_alt}) was generated 
using $q_{ij}=\eta_i/(\eta_j(1-\delta)+\eta_i \delta)$. 
%***DAN: Jasmin, I believe the code does not produce the alternate-qij plot yet, but eventually it needs to.
Though results differed in detail, main conclusions were the same.
%***DAN: make sure this is indeed the case once the plot is generated.
}\label{fig:fig4}
\end{figure}

>>>>

%the beta fucundities version
The \emph{beta fecundities lottery model}
used $B_i = \eta_i F_\beta^{-1} \circ \varphi(b_i)$, where $\varphi$ is the CDF of the standard normal
distribution and $F_\beta$ is the CDF of a beta distribution 
with shape and scale parameters $0.5$ (probability density function proportional to $x^{-05}(1-x)^{-0.5}$ on
the unit interval). So the beta fecundities model had parameters $\eta_1$, $\eta_2$ and $\delta$,
and produced fecundities with $U$-shaped distributions between $0$ and $\eta_1$, for $B_1$, and
$\eta_2$, for $B_2$. The mean fecundity for species $i$ was $\eta_i/2$, and larger $\eta_i$ also 
corresponded to more variable fecundities for species $i$. The parameter $\eta_i$ represents the upper
bound on what fecundities could occur for species $i$.

@@

For each combination of $\delta=0.2, 0.4, 0.6$ and $\mu_1-\mu_2=0,-0.5,-2,-4$,
we plotted the components of (\ref{eq:full_decomp}) against $\sigma$ for $\sigma$ ranging 
from $0$ to $7$. \cite{Chesson_1981} showed that coexistence is impossible for $\delta=1$, and it
is known that the storage effect (and therefore its ATA component) is weak for larger $\delta$, so we 
only considered values of $\delta$ up to $0.6$. Some of 
the $\mu$ differences and $\sigma$ values we considered were very unrealistic (see Results),
so we considered after plotting whether conclusions about the importance of ATAs depended 
on realistic values of parameters.
The simulation methods of \cite{Ellner_2019} were used to compute the components of (\ref{eq:full_decomp}).
Computationally efficient means of performing those simulations are in 
SI section \ref{SI-sect:efficient_computation}.

@@

For each combination of $\delta=0.2, 0.4, 0.6$
and for left- and right-tail associated noise, we plotted 
the components of (\ref{eq:full_decomp}) against the upper bound ratio $\eta_2/\eta_1$, using 
the fairly realistic (see Results) range $1 \leq \eta_2/\eta_1 \leq 5$. 
The simulation methods of \cite{Ellner_2019} were again used.
Computationally efficient means of performing these simulations are in 
SI section \ref{SI-sect:efficient_computation}.

@@

%G2 
To begin fulfilling goal G2 of the Introduction (to explore, using the lottery model, the relative magnitudes 
of the contributions to coexistence that come from ATAs compared to other mechanisms; and to see if ATA rescue 
or ATA exclusion can occur), Fig.
\ref{fig:fig3} shows that, for the log-normal fecundities lottery model, 
the ATA term of (\ref{eq:full_decomp}) sometimes, but not always, has 
magnitude comparable to the magnitudes of other terms of the extended MCT decomposition. 
Magnitudes of $\Delta_i^{[EC]}$ tended to be larger, relative to the magnitudes of the 
other terms of (\ref{eq:full_decomp}), for smaller-magnitude differences between $\mu_1$ and 
$\mu_2$, i.e., for closer-to-equal competitors. 
In the $\mu_1=\mu_2$ case (panels a-c), for which the two species were equal 
competitors, storage effects ($\Delta_i^{[EC]}+\Delta_i^{[E\nat C]}$) 
were the only means by which coexistence could occur, and ATA effects
tended to be negative, weakening total storage effects.

The ATA term 
was also often, or always (for the parameters we considered), comparable in magnitude to the other terms of 
the decomposition (\ref{eq:full_decomp}) for the beta fecundities lottery model 
(Fig. \ref{fig:fig4}). The term contributed negatively to coexistence for environmental noise/fecundities
showing left-tailed association, and contributed positively for noise/fecundities showing right-tailed 
association.

Our results also show that ATAs can make the difference between coexistence 
and failure of two species to coexist: both ``ATA rescue'' and ``ATA
exclusion'' (see Theory) are possible. Starting with the log-normal fecundities 
lottery model, for some of the panels on Fig. \ref{fig:fig3}, the GWR line falls between the ATA
line and the $x$ axis for some values of $\sigma$, e.g., see panel i.
Such cases correspond to parameter combinations
for which the presence or absence of ATAs determine whether the species coexist or not. In the 
case of panel i, for values of $\sigma$ around $6$, 
species 1 is excluded with ATAs but the two species can coexist without ATAs. 
Fig. \ref{fig:fig4} highlights additional examples using the beta fecundities lottery model; 
examples were more common than for the log-normal fecundities model.

For the log-normal fecundities lottery model, the ATA term from (\ref{eq:full_decomp}) tended
only to be comparable in magnitude to the other terms, and ATA exclusion or rescue tended only to
occur, for unrealistic parameters values; but ATA terms were large and ATA rescue and exclusion occurred
for many realistic parameter values for the beta fecundities lottery model. For instance, the ATA term in 
Fig. \ref{fig:fig3} had fairly 
small magnitude, compared to the other terms of (\ref{eq:full_decomp}), for values of $\sigma$ less than 
about $3$. And ATA rescue and exclusion only occurred for $\sigma$ greater than about $3$.
The standard deviation of fecundity for species 1 when 
$\mu_1=0$ and $\sigma=3$ was greater than $8000$, so values of $\sigma$ greater than $3$ are 
probably unrealistic. Thus ATAs seem to be unimportant for coexistence
for the log-normal fecundities lottery model with realistic parameter values. However, 
all the $\eta_2/\eta_1$ ratio values we plotted were fairly realistic, corresponding to
upper-bound fecundities that were up to $5$ times higher for species 2 than for species 1. And 
the ATA term plotted in Fig. \ref{fig:fig4} was comparable in magnitude 
to the other components of (\ref{eq:full_decomp}) across the whole range of parameters
considered. And both ATA rescue and exclusion occurred for values of $\eta_2/\eta_1$ between $1$ and $2$.
Apparently the question of whether ATAs are important for realistic parameter values depends on specifics.

@@

%***DAN: Jasmin, this figure looks really great. Here are a few minor things I would like to ask you to
%do to further optimize:
%1) Currently there are axis labels and numbers only on the bottom and left panels, which makes sense.
%But the axis tick marks needs to occur on all panels. I noticed that this would be an imporvement
%when I found myself wanting to see where the ATA line was between the x axis and the GWR line, but it was 
%a little hard to see where the x axis was becuase there were no tick marks. It is possible to have tick marks
%but no numbers of labels in R, and that's what we need here.
%2) The font of the x axis label, "log fcundity standard deviation, \sigma" seems slightly different from the 
%other textual labels. Can you please make it consistent? Perhaps it is not bold and the others are bold, or 
%something.
\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLLN.pdf}
\caption{The contribution of ATAs to species coexistence was sometimes, but not always, 
meaningfully large compared to 
other components of the MCT decomposition for the log-normal fecundities lottery model.
All panels show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). For each panel, 
$\delta$ and $\mu_1-\mu_2$ 
are fixed at values specified in the margins, and GWR and components of the decomposition
are plotted against $\sigma$. The contribution of ATAs ($\Delta_i^{[EC]}$) sometimes, but not always, 
has magnitude comparable to other components, indicating that ATAs can contribute meaningfully
to species coexistence or failure to coexist. In some instances, the GWR line falls 
between the ATA line and the $x$ axis, indicating that, for those parameter values, ATAs 
make the difference between species
coexistence and exclusion of the weaker competitor. These instances, which are shaded, correspond to
ATA rescue or exclusion (Theory). Standard errors of plotted quantities 
were never greater than \Sexpr{round(se3,5)}, so are not displayed. This plot uses $q_{ij}=1$. An 
analogous plot (SI Fig. \ref{SI-fig:fig3_alt}) uses 
$q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$; though results differed
in detail, main conclusions were the same. \label{fig:fig3}}
\end{figure}

%***DAN: Jasmin, panel labels(a, b, c, etc.) need to be added. Please use the same format for these labels
%as you do in the other figures.
\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/decompFigLB.pdf}
\caption{The contribution of ATAs to species coexistence was often meaningfully large compared to 
other components of the MCT decomposition for the beta fecundities lottery model.
All panels show GWR and the decomposition of
GWR provided by the extended MCT, equation (\ref{eq:full_decomp}). For each panel,
$\delta$ is fixed at the value specified in the upper margin, 
and GWR and components of the decomposition
are plotted against $\eta_2/\eta_1$. The contribution of ATAs ($\Delta_i^{[EC]}$) frequently 
has magnitude comparable to other components, indicating that ATAs can contribute meaningfully
to species coexistence or failure to coexist. In some instances, the GWR line falls 
between the ATA line and the $x$ axis, indicating that, for those parameter values, ATAs make 
the difference between species
coexistence and exclusion of the weaker competitor. These instances, which are shaded, correspond to
ATA rescue or exclusion (Theory). Standard errors of plotted quantities 
were never greater than \Sexpr{round(se4,5)}, so are not displayed. 
This plot uses $q_{ij}=1$. 
%***DAN: Right, Jasmin?
An analogous plot (SI Fig. \ref{SI-fig:fig4_alt}) uses 
$q_{ij}=\eta_i/(\eta_j(1-\delta)+\eta_i \delta)$; 
%***DAN: Jasmin, I believe the code does not produce the alternate-qij plot yet, but eventually it needs to.
though results differed in detail, main conclusions were the same.
%***DAN: make sure this is indeed the case once the plot is generated.
}\label{fig:fig4}
\end{figure}
